### 2023-07-05 paper/localpsf.tex
more edits to numerical results section. Update Hartland23 reference to published version

	We demonstrate the effectiveness of the proposed method by using it to build preconditioners for the Hessian operator arising in two inverse problems governed by partial differential equations (PDEs): inversion for the basal friction coefficient in an ice sheet flow problem and inversion for the initial condition in a diffusive transport problem.
	While for many ill-posed inverse problems the Hessian of the data misfit term exhibits a low rank structure, and hence a low rank approximation is suitable, for many problems of practical interest the numerical rank of the Hessian is still large. This makes conventional matrix-free Hessian approximation methods expensive in terms of the required number of Hessian operator applications (hence PDE solves). In contrast, the performance of the proposed method depends on the locality and shapes of the impulse responses rather than the numerical rank. Hessian impulse responses typically become more local as the numerical rank increases, which makes high rank Hessians perfect candidates for the proposed approach. 
%	To demonstrate the effectiveness of the approximation, w
%	We use the proposed approach to build preconditioners to accelerate the solution of the inverse problems. 
%	We compare the results with classical regularization-based preconditioning and no preconditioning. 
%	The numerical results demonstrate the superior performance (measured in number of PDE solves) of the inverse solver when
	We observe that using the proposed PSF-based preconditioner clusters the spectrum of the preconditioned Hessian near one, yielding roughly 5X-10X reductions in the number of PDE solves required to solve the inverse problems, as compared to classical regularization-based preconditioning and no preconditioning. Finally, we present a comprehensive numerical study for the influence of the diffusion parameters and simulation time (for the diffusive transport problem) on the effectiveness of the approximation. These parameters have an effect on the shape of the impulse responses and on the rank of the Hessian. The results reveal that PSF-based preconditioners are able to form good approximations of high-rank Hessians using a small number of operator applications.
\end{abstract}

@@

Examples of inverse problems with high rank data misfit Hessians include large-scale ice sheet inverse problems \cite{HartlandEtAl23,IsaacEtAl15}, advection dominated advection-diffusion inverse problems~\cite{AkcelikBirosDraganescuEtAl05}\cite[Chapter 5]{Flath13}, high frequency wave propagation inverse problems~\cite{BuiEtAl13}, inverse problems governed by high Reynolds number flows, and more generally, all inverse problems in which the observations highly inform the parameter. The eigenvalues of the data misfit Hessian characterize how informative the data are about components of the parameter in the corresponding eigenvector directions,
hence more informative data leads to larger eigenvalues and a larger numerical rank~\cite{AlexanderianGloorGhattas16}\cite[Section 1.4 and Chapter 4]{Alger19}. Roughly speaking, the numerical rank of the data misfit Hessian is the dimension of the subspace of parameter space that is informed by the data. The numerical rank of the regularization preconditioned data misfit Hessian may be reduced by increasing the strength of the regularization, but this throws away useful information: components of the parameter that could be learned from the observations
would instead be reconstructed based on the regularization~\cite[Section 4]{AlgerEtAl17}\cite[Chapters 1 and 7]{Vogel02}. Hence, low rank approximation methods suffer from a predicament: if the data highly inform the parameter and the regularization is chosen appropriately, then a large number of operator applications are required to form an accurate approximation of the Hessian using low rank approximation methods. 
High rank Hessian approximation methods are thus needed.

Recently there have been improvements in matrix-free H-matrix construction methods in which an operator is applied to structured random vectors, and the response of the operator to those random vectors is processed to construct an H-matrix approximation~\cite{LevittMartinsson22,LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}. 
These methods (which we do not use here) have been used to approximate Hessians in PDE constrained inverse problems~\cite{AmbartsumyanEtAl20,HartlandEtAl23}. 
Although these methods are promising,
the required number of operator applications is still large (e.g., hundreds to thousands).
In this paper, we also form an H-matrix Hessian approximation. However, to reduce the required number of operator applications, we first form a PSF approximation of the data misfit Hessian by exploiting locality and non-negative integral kernel properties, then form the H-matrix using classical H-matrix techniques. By using this two stage approach, we reduce the number of operator applications to a few dozen at most. One limitation of our method is that not all data misfit Hessians satisfy the local non-negative integral kernel properties (for example, the wave inverse problem data misfit Hessian has a substantial amount of negative entries in its integral kernel). However, many data misfit Hessians of practical interest do satisfy these properties (either exactly or approximately), and the proposed method is targeted at approximating these Hessians.

@@

For this example, we consider a sheet of ice flowing down a mountain (see Figure~\ref{fig:ice_mountain_mesh}). Given observations of the tangential component of the ice velocity on the top surface of the ice, we invert for the logarithm of the unknown spatially varying basal friction Robin coefficient field, which governs the resistance to sliding along the base of the ice sheet. 
The setup, which we briefly summarize, follows~\cite{IsaacEtAl15,PetraEtAl12}.
The region of ice is denoted by
$\icedomain \subset\mathbb{R}^{3}$. The basal, lateral and top parts of the boundary $\partial \icedomain$ are denoted by $\Gamma_{b}$, $\Gamma_{l}$, and $\Gamma_{t}$, respectively.
The governing equations are the linear incompressible Stokes
equations,
%
%
\begin{subequations}\label{Stokeseqn}
	\begin{align}
	%-\nabla \cdot \stress(v, p)
	%&=\bodyforce \,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:1}\\
	%\nabla\cdot \velocity
	  %&=0\,\,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:2}\\
          -\nabla \cdot \stress(v, p)=\bodyforce \text{ and }
          \nabla\cdot \velocity &=0 \quad \text{ in }\icedomain, \label{Stokeseqn:1}\\
	\stress(v, p)\normalvec
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{t}, \label{Stokeseqn:3}\\
	\velocity \cdot\normalvec =0 \text{ and } \tangentop\left(\stress(v, p)\normalvec+\exp\left(\basalfriction\right)\velocity\right)
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{b},\label{Stokeseqn:4} \\
	\stress(v, p)\normalvec+\stokesrobincoeff \velocity &=0\,\,\,\,\,\,\text{ on }\Gamma_{l}.\label{Stokeseqn:5}	
	\end{align}
\end{subequations}
%
%
The solution to these equations is the pair $(\velocity, \pressure)$, where $\velocity$ is the ice flow velocity field\footnote{We do not use bold to denote vector or tensor fields to avoid confusion with vectors that arise from finite element discretizations, which are already denoted with bold.}, and $\pressure$ is the pressure field. Here, $\basalfriction$ is the unknown logarithmic
basal friction field (large $\basalfriction$ corresponds to high resistance to sliding) defined on the surface $\Gamma_b$. The quantity $\bodyforce$ is the body force density due to gravity, $\stokesrobincoeff=10^6$ is a Robin boundary condition constant,  $\normalvec$ is the outward normal and $\tangentop$ is the tangential projection operator that restricts a vector field to its tangential component along the boundary. 
We employ Glen's flow law~\cite{Glen1955}, $\stress(v, p)= 2\eta
\dot{\strain}(v) -I\pressure$,
which is a constitutive law for ice that relates the stress
tensor, $\stress$, to the strain rate tensor,
$\dot{\strain}(v)= \frac{1}{2}\left(
\nabla\velocity+\nabla\velocity^{\top} \right)$. Here $\eta$ is the effective viscosity and $I$ is the unit matrix. Glen's exponent has been chosen to be one, which provides for a linear Stokes model. Note that while the PDE is linear, the parameter-to-solution map, $\basalfriction \mapsto (\velocity, \pressure)$, is nonlinear.

The pressure, $\pressure$, is discretized with first order scalar continuous Galerkin finite elements defined on a mesh of tetrahedra. The velocity, $\velocity$, is discretized with second order continuous Galerkin finite elements on the same mesh. The parameter $\basalfriction$ is discretized with first order scalar continuous Galerkin finite elements on the mesh of triangles that results from restricting the tetrahedral mesh to the basal boundary, $\Gamma_b$. Note that $\Gamma_b$ is a two-dimensional surface embedded in three dimensions due to the mountain topography. We also generate a flattened version of $\Gamma_b$, denoted by $\Omega \subset \mathbb{R}^2$, by ignoring the height coordinate. The parameter $\basalfriction$ is viewed as a function on $\Gamma_b$ for the purpose of solving the Stokes equations, and as a function on $\Omega$ for the purpose of building Hessian approximations and defining the regularization. 
%This flattening is necessary because our operator approximation method requires adding points on this surface together to get other points on the surface, and this is only well-defined in flat space. 
The observations are generated by adding multiplicative
Gaussian noise to the tangential component of the velocity field
restricted to the top surface of the geometry. We use $5\%$ noise in all cases, except for Figure~\ref{fig:stokes_reconstructions} and Table \ref{tab:condition_number} where the noise is varied from $1\%$ to $25\%$ and the regularization is determined by the Morozov discrepancy principle for each noise level. The true basal friction
coefficient and velocity fields, which are obtained by
solving~\eqref{Stokeseqn}, are shown in Figure~\ref{fig:true_beta_u}.

@@

Table~\ref{tab:newton_convergence_table} shows the performance of the preconditioner for accelerating the solution of the optimization problem to reconstruct $\mathbf{q}$ from observations with $5\%$ noise. We build the PSF (5) preconditioner in the third Gauss-Newton iteration, and reuse it for all subsequent Gauss-Newton and Newton iterations. No preconditioning is used in the Gauss-Newton iterations before the PSF (5) preconditioner is built. We compare the proposed method with the most commonly used existing preconditioners: no preconditioning (NONE), and preconditioning by the regularization term in the Hessian (REG). The results show that using PSF (5) reduces the total number of Stokes PDE solves substantially compared to regularization preconditioning and no preconditioning. In Figure~\ref{fig:stokes_reconstructions} we show reconstructions for $1\%$, $5\%$, and $25\%$ noise.
%
%
\begin{figure}
	\begin{subfigure}{0.52\textwidth}
		\centering
		\includegraphics[scale=1.0]{stokes_pcg_convergence.pdf}
	\end{subfigure}
	\begin{subfigure}{0.47\textwidth}
	\begin{center}
		\includegraphics[scale=1.0]{stokes_geigs_figure.pdf}
	\end{center}
    \end{subfigure}
	\caption{(Ice sheet) Left: convergence history for solving $\mathbf{H} \mathbf{x} = \mathbf{b}$ using PCG, where $\mathbf{b}$ has i.i.d. random entries drawn from the standard Gaussian distribution and $\mathbf{H}$ is evaluated at the solution of the inverse problem. 
	Results in these figures are shown for the PSF-based preconditioners with 1, 5, and 25 batches (PSF (1), PSF (5), and PSF(25), respectively), regularization preconditioning (REG), and no preconditioning (NONE). The preconditioner is constructed using $\mathbf{H}_\text{gn}$.
	Right: generalized eigenvalues for generalized eigenvalue problem $\mathbf{H} \mathbf{u}_{k}=\lambda_{k} \mathbf{\preconditioner} \mathbf{u}_{k}$. Here $\mathbf{H}$ is the Hessian and $\mathbf{\preconditioner}$ is the PSF approximation constructed using the Gauss-Newton Hessian, $\mathbf{H}_\text{gn}$, for 1, 5, and 25 batches (PSF (1), PSF (5), and PSF (25), respectively), the regularization Hessian (REG), or the identity matrix (NONE).}
	\label{fig:krylov_convergence}
\end{figure} 
%

@@

In Table~\ref{tab:condition_number}, we show the condition number of the preconditioned Hessian for a variety of noise levels ranging from $1\%$ to $25\%$. Note that the condition number using PSF-based preconditioners is extremely small (ranging between 1 and 10) and relatively stable over this range of noise levels. As expected, PSF (25) outperforms than PSF (5), which outperforms than PSF (1). All PSF-based preconditioners outperform no preconditioning and regularization preconditioning for all noise levels.

@@

\subsection{Example 2: Inversion for the initial condition in an diffusive transport problem}
\label{sec:adv}

Here we consider a time-dependent advection-diffusion equation in
which we seek to infer the unknown spatially varying initial condition, $\ipar$, from noisy observations of the state  
at a final time, $T$. This PDE
models diffusive transport in a domain $\advdomain \subset \R^d$, which is
depicted in Figure~\ref{fig:adv_inversion1}. In this case, the state, $\concentration(x,t)$, could be interpreted as the concentration of a contaminant. The problem description
below closely follows the one in~\cite{PetraStadler11}. The domain
boundaries $\partial \advdomain$ include the outer boundaries as well as the
internal boundaries of the rectangles, which represent buildings.  The
parameter-to-observable map $\iFF$ in this case maps an initial
condition $\ipar \in L^2(\advdomain)$ to observations of the concentration at a final time, $\concentration(x, T)$,  through solution of the advection-diffusion
equation given by
\begin{equation}\label{eq:ad}
  \begin{aligned}
    \concentration_t - \kappa\Delta \concentration + v\cdot\nabla \concentration &= 0 & \quad&\text{in
    }\advdomain\times (0,T), \\
    \concentration(\cdot, 0) &= \ipar  &&\text{in } \advdomain , \\
    \kappa\nabla \concentration\cdot \nu &= 0 &&\text{on } \partial \advdomain \times (0,T).
  \end{aligned}
\end{equation}
Here, $\kappa > 0$ is a diffusivity coefficient, $\nu$ is the boundary unit normal vector, and $T > 0$ is the
final time.  The velocity field, $v:\advdomain \rightarrow \mathbb{R}^d$, is computed by solving the
steady-state Navier-Stokes equations for a two dimensional flow with
Reynolds number 50, with boundary conditions $v(x)=(0,1)$ on the left boundary, $v(x)=(0,-1)$ on the right boundary, and $v(x)=(0,0)$ on the top and bottom boundaries, as in~\cite[Section 3]{PetraStadler11}. We use a checkerboard image for the initial condition (Figure~\ref{fig:adv_initial_condition}) and add $5\%$ multiplicative noise to generate synthetic observations at the final time, $T$. The initial condition, velocity field, noisy observations, and reconstructed initial condition are shown in Figure~\ref{fig:adv_inversion1}. We use $\kappa=3.2$e${-1}$ and $T=1.0$ for all results, except for Table \ref{tab:adv_convergence} and Figure~\ref{tbl:adv_vary_kappa_tf} where we vary $\kappa$ and $T$.

In Table~\ref{tab:adv_convergence} we show the number of PCG iterations, $j$, required to solve the Newton linear system to a relative error tolerance of $\|\widehat{\mathbf{q}} - \widehat{\mathbf{q}}_j\| < 10^{-6} \|\widehat{\mathbf{q}}\|$. The solution of the Newton system that we compare to, $\widehat{\mathbf{q}}$, is found via another PCG iteration with a relative residual tolerance of $10^{-11}$. We show results for $T$ ranging from $0.5$ to $2.0$ and $\kappa$ ranging from $10^{-4}$ to $10^{-3}$ using the PSF-based preconditioners with $1$, $5$, and $25$ batches, regularization preconditioning, and no preconditioning. The results show that PSF-based preconditioning outperforms regularization preconditioning and no preconditioning, with more batches yielding better results.

>>>>

	We demonstrate the effectiveness of the proposed method by using it to build preconditioners for the Hessian operator arising in two inverse problems governed by partial differential equations (PDEs): inversion for the basal friction coefficient in an ice sheet flow problem and inversion for the initial condition in a diffusive transport problem.
	While for many ill-posed inverse problems the Hessian of the data misfit term exhibits a low rank structure, and hence a low rank approximation is suitable, for many problems of practical interest the numerical rank of the Hessian is still large. This makes conventional matrix-free Hessian approximation methods expensive in terms of the required number of Hessian operator applications (hence PDE solves). In contrast, the performance of the proposed method depends on the locality and shapes of the impulse responses rather than the numerical rank. Hessian impulse responses typically become more local as the numerical rank increases, which makes high rank Hessians perfect candidates for the proposed approach. 
	We observe that using the proposed PSF-based preconditioner clusters the spectrum of the preconditioned Hessian near one, yielding roughly 5X--10X reductions in the number of PDE solves required to solve the inverse problems, as compared to classical regularization-based preconditioning and no preconditioning. Finally, we present a comprehensive numerical study for the influence of the diffusion parameters and simulation time (for the diffusive transport problem) on the effectiveness of the approximation. These parameters have an effect on the shape of the impulse responses and on the rank of the Hessian. The results reveal that PSF-based preconditioners are able to form good approximations of high-rank Hessians using a small number of operator applications.
\end{abstract}

@@

Examples of inverse problems with high rank data misfit Hessians include large-scale ice sheet inverse problems \cite{Hartland_2023,IsaacEtAl15}, advection dominated advection-diffusion inverse problems~\cite{AkcelikBirosDraganescuEtAl05}\cite[Chapter 5]{Flath13}, high frequency wave propagation inverse problems~\cite{BuiEtAl13}, inverse problems governed by high Reynolds number flows, and more generally, all inverse problems in which the observations highly inform the parameter. The eigenvalues of the data misfit Hessian characterize how informative the data are about components of the parameter in the corresponding eigenvector directions,
hence more informative data leads to larger eigenvalues and a larger numerical rank~\cite{AlexanderianGloorGhattas16}\cite[Section 1.4 and Chapter 4]{Alger19}. Roughly speaking, the numerical rank of the data misfit Hessian is the dimension of the subspace of parameter space that is informed by the data. The numerical rank of the regularization preconditioned data misfit Hessian may be reduced by increasing the strength of the regularization, but this throws away useful information: components of the parameter that could be learned from the observations
would instead be reconstructed based on the regularization~\cite[Section 4]{AlgerEtAl17}\cite[Chapters 1 and 7]{Vogel02}. Hence, low rank approximation methods suffer from a predicament: if the data highly inform the parameter and the regularization is chosen appropriately, then a large number of operator applications are required to form an accurate approximation of the Hessian using low rank approximation methods. 
High rank Hessian approximation methods are thus needed.

Recently there have been improvements in matrix-free H-matrix construction methods in which an operator is applied to structured random vectors, and the response of the operator to those random vectors is processed to construct an H-matrix approximation~\cite{LevittMartinsson22,LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}. 
These methods (which we do not use here) have been used to approximate Hessians in PDE constrained inverse problems~\cite{AmbartsumyanEtAl20,Hartland_2023}. 
Although these methods are promising,
the required number of operator applications is still large (e.g., hundreds to thousands).
In this paper, we also form an H-matrix Hessian approximation. However, to reduce the required number of operator applications, we first form a PSF approximation of the data misfit Hessian by exploiting locality and non-negative integral kernel properties, then form the H-matrix using classical H-matrix techniques. By using this two stage approach, we reduce the number of operator applications to a few dozen at most. One limitation of our method is that not all data misfit Hessians satisfy the local non-negative integral kernel properties (for example, the wave inverse problem data misfit Hessian has a substantial amount of negative entries in its integral kernel). However, many data misfit Hessians of practical interest do satisfy these properties (either exactly or approximately), and the proposed method is targeted at approximating these Hessians.

@@

For this example, we consider a sheet of ice flowing down a mountain (see Figure~\ref{fig:ice_mountain_mesh}). Given observations of the tangential component of the ice velocity on the top surface of the ice, we invert for the logarithm of the unknown spatially varying basal friction Robin coefficient field, which governs the resistance to sliding along the base of the ice sheet. 
The setup, which we briefly summarize, follows~\cite{IsaacEtAl15,PetraEtAl12}.
The region of ice is denoted by
$\icedomain \subset\mathbb{R}^{3}$. The basal, lateral and top parts of the boundary $\partial \icedomain$ are denoted by $\Gamma_{b}$, $\Gamma_{l}$, and $\Gamma_{t}$, respectively.
The governing equations are the linear incompressible Stokes
equations,
%
%
\begin{subequations}\label{Stokeseqn}
	\begin{align}
	%-\nabla \cdot \stress(v, p)
	%&=\bodyforce \,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:1}\\
	%\nabla\cdot \velocity
	  %&=0\,\,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:2}\\
          -\nabla \cdot \stress(v, p)=\bodyforce \text{ and }
          \nabla\cdot \velocity &=0 \quad \text{ in }\icedomain, \label{Stokeseqn:1}\\
	\stress(v, p)\normalvec
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{t}, \label{Stokeseqn:3}\\
	\velocity \cdot\normalvec =0 \text{ and } \tangentop\left(\stress(v, p)\normalvec+\exp\left(\basalfriction\right)\velocity\right)
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{b},\label{Stokeseqn:4} \\
	\stress(v, p)\normalvec+\stokesrobincoeff \velocity &=0\,\,\,\,\,\,\text{ on }\Gamma_{l}.\label{Stokeseqn:5}	
	\end{align}
\end{subequations}
%
%
The solution to these equations is the pair $(\velocity, \pressure)$, where $\velocity$ is the ice flow velocity field\footnote{We do not use bold to denote vector or tensor fields to avoid confusion with vectors that arise from finite element discretizations, which are already denoted with bold.}, and $\pressure$ is the pressure field. Here, $\basalfriction$ is the unknown logarithmic
basal friction field (large $\basalfriction$ corresponds to large resistance to sliding) defined on the surface $\Gamma_b$. The quantity $\bodyforce$ is the body force density due to gravity, $\stokesrobincoeff=10^6$ is a Robin boundary condition constant,  $\normalvec$ is the outward unit normal and $\tangentop$ is the tangential projection operator that restricts a vector field to its tangential component along the boundary. 
We employ a Newtonian constitutive law, $\stress(v, p)= 2\eta
\dot{\strain}(v) -I\pressure$,
where $\stress$ is the stress tensor and
$\dot{\strain}(v)= \frac{1}{2}\left(
\nabla\velocity+\nabla\velocity^{\top} \right)$ is the strain rate tensor~\cite{Glen1955}. Here $\eta$ is the viscosity and $I$ is the identity operator. 
%Glen's exponent has been chosen to be one, which provides for a linear Stokes model. 
Note that while the PDE is linear, the parameter-to-solution map, $\basalfriction \mapsto (\velocity, \pressure)$, is nonlinear.

The pressure, $\pressure$, is discretized with first order scalar continuous Galerkin finite elements defined on a mesh of tetrahedra. The velocity, $\velocity$, is discretized with second order continuous Galerkin finite elements on the same mesh. The parameter $\basalfriction$ is discretized with first order scalar continuous Galerkin finite elements on the mesh of triangles that results from restricting the tetrahedral mesh to the basal boundary, $\Gamma_b$. Note that $\Gamma_b$ is a two-dimensional surface embedded in three dimensions due to the mountain topography. We also generate a flattened version of $\Gamma_b$, denoted by $\Omega \subset \mathbb{R}^2$, by ignoring the height coordinate\footnote{The proposed PSF method requires translating impulse responses. This requires either a flat domain, or a notion of local parallel transport.}. The parameter $\basalfriction$ is viewed as a function on $\Gamma_b$ for the purpose of solving the Stokes equations, and as a function on $\Omega$ for the purpose of building Hessian approximations and defining the regularization. 
%This flattening is necessary because our operator approximation method requires adding points on this surface together to get other points on the surface, and this is only well-defined in flat space. 
The observations are generated by adding multiplicative
Gaussian noise to the tangential component of the velocity field
restricted to the top surface of the geometry. We use $5\%$ noise in all cases, except for Figure~\ref{fig:stokes_reconstructions} and Table \ref{tab:condition_number} where the noise is varied from $1\%$ to $25\%$ and the regularization is determined by the Morozov discrepancy principle for each noise level. The true basal friction
coefficient and resulting velocity fields, which are obtained by
solving~\eqref{Stokeseqn}, are shown in Figure~\ref{fig:true_beta_u}.

@@

Table~\ref{tab:newton_convergence_table} shows the performance of the preconditioner for accelerating the solution of the optimization problem to reconstruct $\mathbf{q}$ from observations with $5\%$ noise. We build the PSF (5) preconditioner in the third Gauss-Newton iteration, and reuse it for all subsequent Gauss-Newton and Newton iterations. No preconditioning is used in the Gauss-Newton iterations before the PSF (5) preconditioner is built. We compare the proposed method with the most commonly used existing preconditioners: no preconditioning (NONE), and preconditioning by the regularization term in the Hessian (REG). The results show that using PSF (5) reduces the total number of Stokes PDE solves to 70, as compared to 908 for regularization preconditioning and 308 for no preconditioning. In Figure~\ref{fig:stokes_reconstructions} we show reconstructions for $1\%$, $5\%$, and $25\%$ noise.
%
%
\begin{figure}
	\begin{subfigure}{0.52\textwidth}
		\centering
		\includegraphics[scale=1.0]{stokes_pcg_convergence.pdf}
	\end{subfigure}
	\begin{subfigure}{0.47\textwidth}
	\begin{center}
		\includegraphics[scale=1.0]{stokes_geigs_figure.pdf}
	\end{center}
    \end{subfigure}
	\caption{(Ice sheet) Left: convergence history for solving $\mathbf{H} \mathbf{x} = \mathbf{b}$ using PCG, where $\mathbf{b}$ has i.i.d. random entries drawn from the standard Gaussian distribution and $\mathbf{H}$ is evaluated at the solution of the inverse problem. 
	Results in these figures are shown for the PSF-based preconditioners with 1, 5, and 25 batches (PSF (1), PSF (5), and PSF(25), respectively), regularization preconditioning (REG), and no preconditioning (NONE). The preconditioner is constructed using $\mathbf{H}_\text{gn}$.
	Right: generalized eigenvalues for generalized eigenvalue problem $\mathbf{H} \mathbf{u}_{k}=\lambda_{k} \mathbf{\preconditioner} \mathbf{u}_{k}$. Here $\mathbf{H}$ is the Hessian and $\mathbf{\preconditioner}$ is the PSF approximation constructed using the Gauss-Newton Hessian, $\mathbf{H}_\text{gn}$, for 1, 5, and 25 batches (PSF (1), PSF (5), and PSF (25), respectively), the regularization Hessian (REG), or the identity matrix (NONE).}
	\label{fig:krylov_convergence}
\end{figure} 
%

@@

In Table~\ref{tab:condition_number}, we show the condition number of the preconditioned Hessian for a variety of noise levels ranging from $1\%$ to $25\%$. Note that the condition number using PSF-based preconditioners is extremely small (ranging between 1 and 10) and relatively stable over this range of noise levels. As expected, PSF (25) outperforms PSF (5), which outperforms PSF (1). All PSF-based preconditioners outperform no preconditioning and regularization preconditioning by several orders of magnitude for all noise levels.

@@

\subsection{Example 2: Inversion for the initial condition in a diffusive transport problem}
\label{sec:adv}

Here we consider a time-dependent advection-diffusion equation in
which we seek to infer the unknown spatially varying initial condition, $\ipar$, from noisy observation of the full state  
at a final time, $T$. This PDE
models diffusive transport in a domain $\advdomain \subset \R^d$, which is
depicted in Figure~\ref{fig:adv_inversion1}. In this case, the state, $\concentration(x,t)$, could be interpreted as the concentration of a contaminant. The problem description
below closely follows the one in~\cite{PetraStadler11}. The domain
boundaries $\partial \advdomain$ include the outer boundaries as well as the
internal boundaries of the rectangles, which represent buildings.  The
parameter-to-observable map $\iFF$ in this case maps an initial
condition $\ipar \in L^2(\advdomain)$ to the concentration field at a final time, $\concentration(x, T)$,  through solution of the advection-diffusion
equation given by
\begin{equation}\label{eq:ad}
  \begin{aligned}
    \concentration_t - \kappa\Delta \concentration + v\cdot\nabla \concentration &= 0 & \quad&\text{in
    }\advdomain\times (0,T), \\
    \concentration(\cdot, 0) &= \ipar  &&\text{in } \advdomain , \\
    \kappa\nabla \concentration\cdot \nu &= 0 &&\text{on } \partial \advdomain \times (0,T).
  \end{aligned}
\end{equation}
Here, $\kappa > 0$ is a diffusivity coefficient, $\nu$ is the boundary unit normal vector, and $T > 0$ is the
final time.  The velocity field, $v:\advdomain \rightarrow \mathbb{R}^d$, is computed by solving the
steady-state Navier-Stokes equations for a two dimensional flow with
Reynolds number 50, with boundary conditions $v(x)=(0,1)$ on the left boundary, $v(x)=(0,-1)$ on the right boundary, and $v(x)=(0,0)$ on the top and bottom boundaries, as in~\cite[Section 3]{PetraStadler11}. We use a checkerboard image for the initial condition (Figure~\ref{fig:adv_initial_condition}) and add $5\%$ multiplicative noise to generate synthetic observations at the final time, $T$. The initial condition, velocity field, noisy observations, and reconstructed initial condition are shown in Figure~\ref{fig:adv_inversion1}. We use $\kappa=3.2$e${-1}$ and $T=1.0$ for all results, except for Table \ref{tab:adv_convergence} and Figure~\ref{tbl:adv_vary_kappa_tf} where we vary $\kappa$ and $T$.

In Table~\ref{tab:adv_convergence} we show the number of PCG iterations, $j$, required to solve the Newton linear system to a relative error tolerance of $\|\widehat{\mathbf{q}} - \widehat{\mathbf{q}}_j\| < 10^{-6} \|\widehat{\mathbf{q}}\|$. The solution of the Newton system to which we compare, $\widehat{\mathbf{q}}$, is found via another PCG iteration with a relative residual tolerance of $10^{-11}$. We show results for $T$ ranging from $0.5$ to $2.0$ and $\kappa$ ranging from $10^{-4}$ to $10^{-3}$ using the PSF-based preconditioners with $1$, $5$, and $25$ batches, regularization preconditioning, and no preconditioning. The results show that PSF-based preconditioning outperforms regularization preconditioning and no preconditioning, with more batches yielding better results.
